Full Title

### 2018-01-27 cones/paper/cones.tex
Finished full abstraction proof.

Let $T$ be an Idealized Algol type, and let $M,N\from T$ be terms.  
Then we write $M\lmam N$ if for all compatible contexts $C[-]$ of ground type we have
\begin{gather*}
  C[M]\converges V \Rightarrow C[N] \converges V\\
  C[M]\mustconverge \Rightarrow C[N] \mustconverge
\end{gather*}

@@

\begin{lemma}
  Let $A,B,C$ be games and let $\sigma\from A\implies B$, $\tau\from B\implies C$ be deterministic complete strategies.  
  Then $\sigma;\tau$ is complete.
\end{lemma}
\begin{proof}
  The proof relies on a lemma from \cite{hoPcf} that states (in our language) that if $\sigma$ and $\tau$ are deterministic strategies and $s\in T_{\sigma;\tau}$ then there is a unique minimal $\s\in T_\sigma\|T_\tau$ such that $\s\vert_{A,C}=s$.  
  That means that if $s_1\prefix s_2\prefix \cdots$ is an infinite increasing sequence of plays in $T_{\sigma;\tau}$, with infinite limit $s$, then there is a corresponding infinite increasing sequence of legal interactions $\s_1 \prefix \s_2 \prefix \cdots$.  
  Then the limit of this sequence is an infinite legal interaction $\s$ and we must have $\s\vert_{A,B}\in\sigma$, $\s\vert_{B,C}\in\tau$ by completeness of $\sigma$ and $\tau$.  
  Therefore, $s=\s\vert_{A,C}\in T_{\sigma;\tau}$.
\end{proof}

\begin{corollary}
  The composition of strategies $\sigma\from A\implies B$ and $\tau\from B \implies C$ is a well-formed strategy for $A\implies C$.
\end{corollary}
\begin{proof}
  The only tricky point is eastablishing that diverge-or-reply holds for $\sigma;\tau$.  
  Again, it is sufficient to prove this in the case that $\sigma$ and $\tau$ are deterministic and complete.
  Then it essentially follows from the argument used in \cite{abramskyjagadeesangames} that shows that a partiality at an $O$-position $s\in T_{\sigma;\tau}$ must arise either from a partiality in $T_\sigma$ or $T_\tau$ or from `infinite chattering' between $\sigma$ and $\tau$.  
  In the first case, the diverge-or-reply rule for $\sigma$ and $\tau$ gives us a divergence at $s$ in $\sigma;\tau$.  
  In the second case, an infinite chattering between $\sigma$ and $\tau$ corresponds to an infinite interaction $\s\in\int(A,B,C)$ ending with infinitely many moves in $B$ such that $\s\vert_{A,C}=s$. 
  Completeness for $\sigma$ and $\tau$ tells us that $\s\vert_{A,B}\in D_\sigma$ and $\s\vert_{B,C}\in D_\tau$ and therefore that $\s\vert_{A,C}\in D_{\sigma;\tau}$.  
\end{proof}

@@

>>>>

If $\ia\emptyset M$ is a configuration with empty store, we call $M$ a \emph{closed term}.  
Let $T$ be an Idealized Algol type, and let $M,N\from T$ be closed terms.
Then we write $M\lmam N$ if for all compatible contexts $C[-]$ of ground type we have
\begin{gather*}
  C[M]\converges V \Rightarrow C[N] \converges V\\
  C[M]\mustconverge \Rightarrow C[N] \mustconverge
\end{gather*}

@@

\begin{lemma}
  Let $A,B,C$ be games and let $\sigma\from A\implies B$, $\tau\from B\implies C$ be deterministic complete strategies.  
  Then $\sigma;\tau$ is complete.
  \label{lem:complete-deterministic-composition}
\end{lemma}
\begin{proof}
  The proof relies on a lemma from \cite{hoPcf} that states (in our language) that if $\sigma$ and $\tau$ are deterministic strategies and $s\in T_{\sigma;\tau}$ then there is a unique minimal $\s\in T_\sigma\|T_\tau$ such that $\s\vert_{A,C}=s$.  
  That means that if $s_1\prefix s_2\prefix \cdots$ is an infinite increasing sequence of plays in $T_{\sigma;\tau}$, with infinite limit $s$, then there is a corresponding infinite increasing sequence of legal interactions $\s_1 \prefix \s_2 \prefix \cdots$.  
  Then the limit of this sequence is an infinite legal interaction $\s$ and we must have $\s\vert_{A,B}\in\sigma$, $\s\vert_{B,C}\in\tau$ by completeness of $\sigma$ and $\tau$.  
  Therefore, $s=\s\vert_{A,C}\in T_{\sigma;\tau}$.
\end{proof}

\begin{corollary}
  The composition of strategies $\sigma\from A\implies B$ and $\tau\from B \implies C$ is a well-formed strategy for $A\implies C$.
  \label{cor:diverge-or-reply}
\end{corollary}
\begin{proof}
  The only tricky point is eastablishing that diverge-or-reply holds for $\sigma;\tau$.  
  Again, it is sufficient to prove this in the case that $\sigma$ and $\tau$ are deterministic and complete.
  Then it essentially follows from the argument used in \cite{abramskyjagadeesangames} that shows that a partiality at an $O$-position $s\in T_{\sigma;\tau}$ must arise either from a partiality in $T_\sigma$ or $T_\tau$ or from `infinite chattering' between $\sigma$ and $\tau$.  
  In the first case, the diverge-or-reply rule for $\sigma$ and $\tau$ gives us a divergence at $s$ in $\sigma;\tau$.  
  In the second case, an infinite chattering between $\sigma$ and $\tau$ corresponds to an infinite interaction $\s\in\int(A,B,C)$ ending with infinitely many moves in $B$ such that $\s\vert_{A,C}=s$. 
  Completeness for $\sigma$ and $\tau$ tells us that $\s\vert_{A,B}\in D_\sigma$ and $\s\vert_{B,C}\in D_\tau$ and therefore that $\s\vert_{A,C}\in D_{\sigma;\tau}$.  
\end{proof}

Our proof for Corollary \ref{cor:diverge-or-reply} really makes use of the fact that a locally complete strategy is \emph{lively} in the sense of \cite{LevyGsInfinite}; i.e., locally deterministic.  
Our definition is slightly stronger than liveliness, because it insists that the union of complete strategies be \emph{countable}.  
This will be essential to our definability result.

@@

\section{Full abstraction}

\subsection{Denotational Semantics}

The category in which we shall model our language is the category $\G_{ND,vis,rec}^\oc$ -- the Cartesian closed category of (enumerated) games with nondeterministic, recursively presentable visible strategies.  
We have a natural embedding $\G_{D,vis,rec}^\oc\hookrightarrow\G_{ND,vis,rec}^\oc$, and we know that $\G_{D,vis,rec}^\oc$ is a universal and fully abstract model of Idealized Algol.

Any term $M\from T$ of Idealized Algol with countable nondeterminism may be written as $M = C[\wn]$, where $C$ is a context not involving the constant $\wn$.  
Then the term $\lambda n.C[n]$ is a term of Idealized Algol, and therefore has a denotation $\oc\bN \to \deno{T}$ as in \cite{SamsonGuyIAPassive}.
We define the denotation of $M$ to be given by the composite
\[
  I \xrightarrow{\top_{\bN}}
  \oc \bN \xrightarrow{\deno{\lambda n.C[n]}}
  \deno{T}
  \]
In other words, we interpret the constant $\wn$ using the strategy $\top_{\bN}$ for $\bN$.

\subsection{Adequacy and soundness}

We prove a \emph{consistency} result for our model.  
Let $M$ be a closed term of type $\com$.

\begin{lemma}
  If $M\converges\skipp$ then $qa\in T_{\deno M}$.  
  If $M\mustconverge$ then $D_{\deno M}=\emptyset$.
  \label{lem:consistency}
\end{lemma}

We can also prove the converse, \emph{computational adequacy}.

\begin{proposition}
  If $qa\in T_{\deno M}$ then $M\converges\skipp$.
  If $D_{\deno M}=\emptyset$ then $M\mustconverge$.
  \label{prop:adequacy}
\end{proposition}

In order to prove these, we come back to the concept of an \emph{evaluation} of a term as a sequence of natural numbers that we use to replace the constant $\wn$.

\begin{lemma}
  Let $M=C[\wn]$ be a term of type $\com$.  
  Write $\sigma_M$ for the denotation of the term $\lambda n.C[n]$.  
  Let $\pi$ be some evaluation of the configuration $\ia s M$, where $s$ is some state.  
  Then there is a deterministic total strategy $\sigma_\pi\from \oc\bN$ such that
  \begin{itemize}
    \item $\pi$ terminates at the value $\skipp$ if and only if then $qa\in T_{\sigma_\pi;\sigma_M}$.
    \item $\pi$ diverges if and only if $D_{\sigma_\pi;\sigma_M}\ne\emptyset$.
  \end{itemize}
  \label{lem:soundness-lemma}
\end{lemma}

\begin{proof}
  Let $n_1,\dots,n_k,d$ be a finite sequence of natural numbers.  
  We define an Idealized Algol term $N_{n_1,\dots,n_k,d}\from (\nat \to \com) \to \com$ to be the following.
  \[
    \lambda f.\new_\nat (\lambda v.f(v\coloneqq (suc\;\deref v); \case_{k+1}\;\deref v\;\Omega\;n_1 \cdots n_k d))\,.
    \]
  Here, $\case_{k+1}\;a\;n_0\;\cdots\;n_k\;d$ evaluates to $n_i$ if $a$ evaluates to $i$, and evaluates to $d$ if $a$ evaluates to $j> k$.
  Clearly, it can be built out of $\IfO$s.
  This term adds an extra variable $v$ to the program; each time $f$ is called, it increments the value of $v$ and maps its value on to one of the $n_i$.

  Now let $\pi$ be a finite evaluation of $\ia s {C[\wn]} \converges\skipp$.  
  Encode $\pi$ as a sequence $n_1,\dots,n_k$.  
  Let $d$ be some arbitrary number.
  Then we can show that the following term also converges to $\skipp$ in the same way:
  \[
    N_{n_1,\dots,n_k,d} (\lambda n.C[n])\,.
    \]
  The idea here is similar to one used in testing; we want to test the behaviour of a nondeterministic program, and to do so we \emph{mock} the random number generator in order to simulate a particular evaluation path using purely deterministic programs.

  If instead $\pi$ is a finite evaluation of $\ia s {C[\wn]}\not\mustconverge$; i.e., a divergent evaluation (that nevertheless only involves finitely many calls to the nondeterministic oracle), then the term $N_{n_1,\dots,n_k,d} (\lambda n.C[n])$ will diverge according to the same execution path.

  Digging into the construction of $\new$ within Idealized Algol, as given in \cite{SamsonGuyIAPassive}, we see that for any term $F$ of type $\nat\to\com$ the denotation of $N_{n_1,\dots,n_k,d} F$ is given by the composite
  \[
    I \xrightarrow{\cell_{0}}
    \oc \Var \xrightarrow{\deno{\lambda v.v\coloneqq (\suc\;\deref v);\case_{k+1}\;\deref v\;\Omega\;n_1\cdots n_k d}}
    \oc\bN \xrightarrow{\deno F}
    \bC\,.
    \]
  We set $\sigma_\pi$ to be the composite of the left two arrows.  
  Observe that $\sigma_\pi$ is the strategy with unique maximal infinite play as follows.
  \[
    q\;n_1\;\cdots\;q\;n_k\;q\;d\;q\;d\;\cdots
    \]
  Setting $F = \lambda n.C[n]$, we see that $\deno{F}=\sigma_M$.  
  So, by soundness and adequacy for the Idealized Algol model, we see that $qa\in T_{\sigma_\pi;\sigma_M}$ if and only if $N_{n_1,\dots,n_k,d} (\lambda n.C[n])\converges\skipp$, which is the case if and only if $M\converges\skipp$ along the evaluation $\pi$.
  Similarly, $D_{\sigma_\pi;\sigma_M}\neq\emptyset$ if and only if $N_{n_1,\dots,n_k,d}(\lambda n.C[n])$ diverges, which is equivalent to saying that $M$ diverges along the evaluation $\pi$.

  Lastly, we need to deal with the case that $\pi$ is an evaluation $n_1,n_2,\dots$ that consults the nondeterministic oracle infinitely often.
  In this case, $M$ must certainly diverge along the evaluation $\pi$.
  In that case, we define $\pi_n^{(j)}$ to be the strategy for $\oc N$ corresponding to the term $N_{n_1,\dots,n_j,\Omega}$.  
  So $\pi_n^{(j)}$ has a unique finite maximal play
  \[
    q\;n_1\;q\;n_2\;\cdots\;q\;n_j\;q\,,
    \]
  at which point the strategy has a partiality.

  Evaluation of the term $N_{n_1,\dots,n_j,\Omega} (\lambda n.C[n])$ must diverge, since it will proceed according to the evaluation $\pi$ and eventually reach the divergence (since $\pi$ consults the oracle infinitely often).  
  This implies that $D_{\sigma_\pi^{(j)};\sigma_M}\ne\emptyset$ for all $j$.

  We define $\sigma_\pi$ to be the least upper bound of the $\sigma_\pi^{(j)}$ (e.g., in the sense of \cite{mcCHFiniteND}).  
  Since composition is continuous for deterministic (!) strategies, we deduce that $D_{\sigma_\pi;\sigma_M}\ne\emptyset$.

  $\sigma_\pi$ has plays of the form
  \[
    q\;n_1\;q\;n_2\;\cdots\,,
    \]
  and so it is total.
\end{proof}

We can now prove our soundness and adequacy results.

\begin{proof}[Proof of Lemma \ref{lem:consistency}]
  Suppose that that $M$ is a closed term of type $\com$ and that $M\converges\skipp$.  
  Let $\pi$ be an evaluation such that $M\converges\skipp$ along $\pi$.  
  Then $qa\in T_{\sigma_\pi;\sigma_M}$ for some strategy $\sigma_\pi$.  
  It follows that $qa\in T_{\top_\bN;\sigma_M}$.

  Now suppose that $D_{\deno M}\neq\emptyset$.  
  Then there is some interaction $\s\in\int_\infty(I,\oc\bN,\bC)$ such that $\s\vert_\bC\in D_{\sigma_M}$.
  Let $\sigma$ be the minimal deterministic strategy for $\oc\bN$ containing the play $\s\vert_{\oc\bN}$.  
  By the proof of Lemma \ref{lem:soundness-lemma}, $\sigma\subset\sigma_\pi$ for some evaluation $\pi$ of $M$.  
  Then we have $D_{\sigma_\pi;\sigma_M}\ne\emptyset$, so $M$ diverges along $\pi$.
\end{proof}

\begin{proof}[Proof of Proposition \ref{prop:adequacy}]
  Suppose that $M$ is a closed term of type $\com$ and that $qa\in T_{\deno M}$.  
  Then there is some interaction $\s\in\int_\infty(I,\oc\bN,\bC)$ such that $qa\in\s\vert_\bC$.  
  Let $\sigma$ be the minimal deterministic strategy for $\oc\bN$ containing the play $\s\vert_{\oc\bN}$.  
  By the proof of Lemma \ref{lem:soundness-lemma}, $\sigma\subset\sigma_\pi$ for some evaluation $\pi$ of the term $M$.
  Then we have $qa\in T_{\sigma_\pi;\sigma_M}$, so $M\converges\skipp$ along $\pi$.

  Now suppse that $M\not\mustconverge$.  
  Let $\pi$ be a divergent evaluation of $M$.  
  Then we have $D_{\sigma_\pi;\sigma_M}\ne\emptyset$.  
  Moreover, since $\sigma_\pi$ is total, it follows that $D_{\top_{\bN};\sigma_M}\ne\emptyset$.
\end{proof}

We define \emph{extensional equivalence of strategies} as follows.  
If $\sigma,\tau$ are two strategies for a game $A$, we say that $\sigma\oes\tau$ if for all test morphisms $\alpha\from A \to \bC$ we have $\sigma;\alpha=\tau;\alpha$.  
Having defined this equivalence, we may prove \emph{soundness} in the usual way.

\begin{theorem}[Soundness]
  Let $M,N$ be two closed terms of type $T$.  
  If $\deno{M}\oes\deno{N}$ then $M\emam N$.  
\end{theorem}

\subsection{Universality}

Let $S,T$ be Idealized Algol types and let $\sigma\from S\to T$ be a morphism in $\G_{ND,vis,rec}^\oc$.  
We want to prove that $\sigma$ is the denotation of some term.

By our nondeterministic factorization result, we know that $\sigma=\top_\bN;\Det(\sigma)$, where $\Det(\sigma)$ is a deterministic strategy.  
By universality for $\G_{D,vis,rec}^\oc$, we know that $\Det(\sigma)=\deno{M}$ for some closed term $M\from S \to T$.  
Then $\sigma=\top_\bN;\Det(\sigma)=\deno{\wn};\deno M = \deno{M\;\wn}$.

\subsection{Full abstraction}

\begin{theorem}[Full abstraction]
  Let $M,N$ be two closed terms of type $T$.  
  If $M\emam N$ then $\deno{M}\oes\deno{N}$.  
\end{theorem}
\begin{proof}
  Let $A=\deno T$.  
  Suppose that $\deno{M}\not{\oes}\deno{N}$; so there is some strategy $\alpha\from A\to\bC$ such that $\deno{M};\alpha\ne\deno{N};\alpha$.  
  By universality, we have $A=\deno{P}$ for some closed term $P$ of type $T\to\com$.  
  Then we have $\deno{M};\deno{P}\ne\deno{N};\deno{P}$; by consistency and adequacy, it follows that $M\not{\emam} N$.
\end{proof}
